---
title: "Student Academic Performance Project Proposal"
subtitle: "Analysis of Factors that contribute to Student Academic Performance"
author: "Lab Group - Rikki Kendall, Tolulemi Gbile, Geary Stonesifer"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA, 
                      message = FALSE, warning = FALSE,
                      fig.height=6, fig.width = 9, fig.align = "center")
```

```{r packages}
library(tidyverse)
library(infer)
```

# Data

```{r load_data}
edu <- read_csv("data/xAPI-Edu-Data.csv")

edu
```

```{r set_seed}
set.seed(5768952)
```

# Section 1 - Introduction

Our general research question is as follows: Is it possible to
predict a child's class level based off of the information provided
in the dataset? If so, which variables are of significance? The
information we're going to be using to unearth the answer to this
research question is from Kaggle.com. The information was collected
from a Learning Management System called Kalboard 360. The data is
collected from a learner activity tracker tool which tracks the
students behavior, like if they're reading an article, watching an
informative video, or doing their classwork. The variables used in
this data set and their data types are as follows:

### Variables 

* Gender - Student's Gender (categorical)
* Nationality - The student's nationality (categorical)
* Place of birth - the student's place of birth (categorical)
* Educational Stage - Elementary, Middle, or High school? (categorical) 
* Grade Level - Grade student is in (categorical)
* Section ID - Classroom the student belongs to (categorical) 
* Topic - Course topic (categorical)
* Semester - School Semester (discrete numerical)
* Parent Responsible for Student (categorical)
* Raised Hand - How many times the student raises their hand in the classroom
(continuous numerical)
* Visited Resources - How many times a student uses the course content 
(continuous numerical)
* Viewing Announcements - How many times the student checks the new 
announcements (continuous numerical)
* Discussion Groups - How many times a student participates in discussion 
groups (continuous numerical)
* Parent Answering Survey - Did the parent answer the surveys provided by the
school? (categorical)
* Parent School Satisfaction - Is the parent satisfied with the school? 
(categorical)
* Student Absence days - How many times the student has missed school 
(categorical)
* Class - Low, Medium, or High, depending on their grades/marks at
the end of the semester 
(categorical) 

# Section 2 - Data analysis plan

The aim of our project is to attempt to predict a child's class based on 
relevant variables provided in the data set. In order to do
so we will be using two different variables: the 'outcome'
(dependent, response, Y), which is the child's class (L,M,H) and the
'predictor' (independent, explanatory, X) which can be any
combination of the aforementioned variables in section 1. 

### Predictor

Explanatory Variables: Any combination of the variables listed above.

### Outcome

Response Variable: Class

### Summary Statistics/Visualizations

In order to attempt to prove that class is predictable, we will run
several different linear regression models. With the given
correlations expected, we should be able to extrapolate a child's
class. However, what is crucial is to pick the most relevant of the
variables. Initially, we will screen the variables based on
intuition as to which are obviously not useful in determining class.
Then we will those that are and run several linear regression
models.

However, at first to portray the relevance of the variables we have
provided several visualizations below to learn more about the data: 

The first graph explores the relationship between the instances of
raised hands and discussion. These two variables, although not
necessarily dependent on one another, intuitively go hand in hand
since both variables are related to class participation.

```{r sec2plot1, fig.width=8, fig.height=6}

ggplot(data = edu, 
       mapping = aes(x = raisedhands, y = Discussion, color = Class)) +
  geom_point() + 
  labs(x = "Raised Hands", y = "Discussion", 
       title = "Visualization of Relationship between Instances of
       Raised Hands and Discussion") + 
  theme_minimal(base_size = 11)

```
Given the disparity in results, whereby lower class is visibly on
the lower spectrum of both Discussion and Raisedhands these
variables will be important. 


The following visualization corroborates the data in the prior
scatter-plot for each country. This is to say that there is a clear
trend among class in various countries given the variable,
'raisedhands.' 

```{r sec3plot2, fig.width = 8, fig.height = 6}
ggplot(edu, aes(x = PlaceofBirth, y = raisedhands, fill = Class)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_classic() + 
  labs(x = "Place of Birth", y = "Raised Hands", 
       title = "Visualization of Relationship between Instances of
       Raised Hands and Place of Birth") + 
  theme_minimal(base_size = 11)
```
To continue to explore the variables, the next scatterplot takes two
more data entries to examine a connection between class and Visited
Resources / Announcements. 

```{r sec2plot3, fig.width=8, fig.height=6}

ggplot(data = edu, 
       mapping = aes(x = VisITedResources, y = AnnouncementsView,
                     color = Class)) + 
  geom_point() + 
  labs(x = "Visited Resources", y = "Announcements", 
       title = "Visualization of Relationship between Visited
       Resources and Announcements") + 
  theme_minimal(base_size = 12)

```
Again, we begin to see some trends when plotting relevant data side
by side. Even more so in this visualization than the first
scatter-plot, there is clear distinction between the classes of
students.

Lastly, I have attempted to draw a connection to class using Parents
satisfaction to examine a potential relationship with class.

```{r sec2plot4}

ggplot(edu, aes(x = ParentschoolSatisfaction, fill = Class)) +
    geom_bar(position = position_dodge()) +
  labs(x = "Parents Satisfaction", y = "Number of Responses", 
       title = "Visualization of Parents School Satisfaction") + 
  theme_minimal(base_size = 12)

```
This data provides interesting insight to the fact that in the
Higher class bracket, there is a significant reduction in instances
of negative feedback to any given school. Conversely, it appears the
lower class bracket more often has negative responses to their
school. 

Given the visualizations of the variables above, it appears there
are significant trends in the data that we aim to use to answer our
research question: Is it possible to predict a child's class level
based off of the information provided in the dataset?

# Section 3 - Data

```{r}
glimpse(edu)
```